---
title: "Confidence Intervals"
author: "Edwin Lo, GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
# knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```


# setup 
Like we did before, let's load our data and create a dataframe call **mlb**, then take a look.

```{r import_data, include=F}
getwd()
# mlb <- data.frame(read.csv("BaseballHeightWeight.csv", header = TRUE))
mlb <- read.csv("BaseballHeightWeight.csv", header = TRUE)
str(mlb)
head(mlb)
colnames(mlb)[4:6]=c("height","weight","age")
# The line above produces no output. To see the changes, use str() or head()
```

Assume we already completed EDA. Let's look at confidence intervals now.


## Confidence Interval / Z-Interval / T-Interval
Let's try these inferential statistics tools that we learned:  
z-interval  
t-interval  

```{r, include=F}
mean.height = mean(mlb$height, na.rm = TRUE); 
mean.weight = mean(mlb$weight, na.rm = TRUE); 
```

Before we do that, let us record again the mean height and weight for the entire mlb population. The mean height is mean.height = `r format(mean.height, digits=5)` inches, mean.weight = `r format(mean.weight, digits=5)` lbs. 

Now let us create a smaller sample of the mlb players out of the entire population, and call it mlbsample.
```{r mlbsample, include = F}
set.seed(321) # just so that everyone have the same sample for comparison
mlbsample = mlb[ sample(nrow(mlb),30), ]
str(mlbsample)
head(mlbsample)
format(mean(mlbsample$height), digits=5)
format(mean(mlbsample$weight, na.rm = TRUE), digits=5)
```

## Z-Interval Estimation (and Z-Test for next class)
t-test and t-interval functions are standard in R, but z-test and z-interval turns out require other libraries. We use `BSDA` here.

```{r z_test1}
loadPkg("BSDA") # for z.test
ztest95 = z.test(x=mlbsample$height, sigma.x = 2.31) # default conf.level = 0.95
ztest95
ztest99 = z.test(x=mlbsample$height, sigma.x = 2.31, conf.level=0.99 )
ztest99
ztest50 = z.test(x=mlbsample$height, sigma.x = 2.31, conf.level=0.50 )
ztest50
```

Let us also take a look at the object "ztest99". It contains these info:
```{r z_test2}
names(ztest99)
ztest99$conf.int
# ztest99$alternative
ztest99$estimate
ztest99$statistic
ztest99$method
# etc etc
```

Now let us compare the three intervals, and note the actual average for the entire population is `r mean.height`.
```{r z_test2b}
mean.height
ztest99$conf.int
ztest95$conf.int
ztest50$conf.int
```

## T-Intervals Estimation (and T-Test for next class)
Similar for t-test, except it is **easier**, since we do not need to know the standard deviation (sigma) of 
the population. First, the results for the default c=95% confidence level.
```{r t_test}
# t.test is included in the basic R package 'stats'
ttest95 = t.test(x=mlbsample$height) # default conf.level = 0.95
ttest95
ttest99 = t.test(x=mlbsample$height, conf.level=0.99 )
ttest99
ttest50 = t.test(x=mlbsample$height, conf.level=0.50 )
ttest50
names(ttest99)
ttest99$conf.int
ttest99$alternative
ttest99$estimate
```


Now let us compare the three intervals, with that produced from the z-test.
```{r t_test2_99}
mean.height
ztest99$conf.int
ttest99$conf.int
```

```{r t_test2_95}
mean.height
ztest95$conf.int
ttest95$conf.int
```

```{r t_test2_50}
mean.height
ttest50$conf.int
ztest50$conf.int
```
You can try the above for the weight data?



