---
title: "R - Assignment Mod 2 Week 3"
author: "Rajkumar Conjeevaram Mohan"
date: "02/03/2022"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
library(ggplot2)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r}
# 1. Answer each question using in words/paragraph. 
# 2. DO NOT use comments inside code blocks (like here) to answer anything. Those are for notes between coders/self. They 
# will be ignored, and not counted as answers. 
# 3. Keep the line/option    knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# You can temporarily turn this on/off and use other option while you are working on the homework if it helps. 
# The submitted work should have this option selected instead.
# 4. All charts/graphs/tables should have appropriate captions. 
```


# HW - Module 2 week 03

## Question 1. Import
**Read in the file with all the baseball players height, weight and age info (BaseballHeightWeight.csv) as a dataframe and save it as baseball. Then display the structure of the dataframe.**  

```{r Q1}
# This is coder's comments
# Since my RMD and csv files are in the same working directory, I do not need to give the full path to the csv file. Alternatively, you can use the functions getwd() and setwd() to handle it.

#baseball <- data.frame(read.csv("BaseballHeightWeight.csv"))
baseball <- read.csv(file = "BaseballHeightWeight.csv",
                     header = T,
                     sep=",",
                     #row.names=1,
                     #fill=T, # I am disabling this as it turns out we are playing with Null values at some point
                     strip.white = T,
                     blank.lines.skip=T,
                     allowEscapes=F)

baseball <- data.frame(baseball)
# baseballdf <- read.csv("BaseballHeightWeight.csv")
# You can compare the two above lines, and look at the object called baseball in the "Environment" on the upper right. 
# Turns out either method gives the same result. In some cases (depending on the .csv file), they might differ. To be safe, I usually use data.frame( ) afterwards.
```

The function `read.csv()` was used to load the dataframe. In addition, we can see that there are `r nrow(baseball)` in the dataframe.
Note that `read.csv()` is just one of the many functions you can use to import data files. 

The structure is printed below:

```{r, result="asis"}
# need results="asis" to overwrite the default, and show the result block in Html/pdf
str(baseball)
```

I have yet to find a simple and easy way to "pretty-print" such structure in R output. Most of the time we don't need to show these to the readers though. 


## Question 2. Head and Tail
**Print out the first 5 and the last 3 rows of the dataframe.**  

The `head()` and `tail()` functions are what we need.

```{r, results='markup'}
# Probably looks more neat
rmarkdown::paged_table(head(baseball, n=5))
rmarkdown::paged_table(tail(baseball, n=3))
```

But we should use the modified version in `ezids` from now on:

```{r, results="asis"}
# you can use "markup" or "asis" here to get the same result.
xkabledplyhead(baseball, 5, "The first 5 rows using `xkabledplyhead` function." )
```
and the last three rows:
```{r, results="asis"}
# you can use "markup" or "asis" here to get the same result.
xkabledplytail(baseball, 3, "The last 3 rows using `xkabledplytail` function." )
```


## Question 3. Header change
**Change the headers/column names from Weight.pounds. to weight, from Height.inches. to height, and Age to age.**  
```{r}
baseball.colnames <- colnames(baseball)
baseball.colnames[c(4,5,6)] <- c("Height", "Weight", "age")
colnames(baseball) <- baseball.colnames
```

## Question 4. Head again
**Print out the first two lines again to see the new column names.**  
```{r, results='asis'}
xkabledplyhead(baseball, 2)
```

## Question 5. Summary
**Find the statistics (summary) of the heights, weights, and ages of the players. Using complete sentences, give the reader these summary.**  
```{r, echo=T, results='asis'}
tmp_s <- xkablesummary(baseball)
print(tmp_s)
# I have given %.2f for age and height instead of %.1f since I can't for now 
# look at every possible value in the list to verify.
```
There are in total 1034 Baseball players with an average age, Weight, and Height of 28.7, 202, 73.7 respectively.  
While most of the player are aged between 25.4 and 31.2, the data shows minimum age of 20.9 and maximum of 48.5  
Also, players were normally found to have a height between 72 and 75, but exceptional case such as being 67 and 83 tall do exist.  
At last, data shows most players weigh between 187 and 202, but there are exceptional cases, also known as outliers from Data Science perspective, that shows weights 150 and 290 as two extremes.


## Question 6. Subset
**You should find an issue with the weight data with an NA value. Use the `subset( )` function, select that data point (save as shyboy) and find out who the shy boy was.**
```{r, results='asis'}
shyboy.index <- which(is.na(baseball[, "Weight"]))
shyboy <- baseball[shyboy.index, ]
sprintf("It turns out the shyboy was %s", shyboy$Name)
```



## Question 7. Remove Shyboy
**Remove shyboy from baseball, and then show the number of observations of the cleaned up dataframe baseball.**  
```{r, results='asis'}
baseball <- baseball[-shyboy.index, ]
print(nrow(baseball))
```


## Question 8. Boxplot
**Make a quick boxplot of the weight of the players. Use the `title( )` function immediately to add a title to the chart. (Try color #0000FF for blue.)**  
```{r}
ggplot(baseball, aes(x='Players', y=Weight)) + 
geom_boxplot(width=0.5, lwd=0.5, outlier.color='red', fill='#0000FF', alpha=0.5) + 
stat_boxplot(geom = "errorbar", # Boxplot with error bars 
                    width = 0.2) +
labs(subtitle="Baseball players weights boxplot") + 
xlab("Aggregated players") + 
ylab("Weight")

```


## Question 9. Histogram
**Make a histogram of the weight of the players. Try color 'green' this time. The function understands the word 'green' just fine. Give it a better title than the default. Does the distribution looks normal?**
```{r}
ggplot(baseball, aes(x=Weight)) + 
# I had to give bins=20 since it was throwing some warning
geom_histogram(fill='green', color='black', alpha=0.3, bins=20) +
labs(subtitle="Baseball players weights boxplot") + 
xlab("Aggregated players") + 
ylab("Weight")
```
The distribution does appear normal.

## Question 10. Scatterplot
**Use ggplot, make a plot with weights vs heights of the players, color by Teams. Give it an appropriate title.**  
```{r}
ggplot(baseball, aes(x=Height, y=Weight)) +
geom_jitter(aes(color=Team), alpha=0.5) +
xlab('Height') +
ylab('Weight') +
labs(subtitle='Comparison of weights against heights of baseball players')
```


## Question 11. Plot by age
**Make a plot with weights vs heights of the players, color by age.**
```{r}
ggplot(baseball, aes(x=Height, y=Weight, col=age)) +  
geom_jitter(alpha=0.5) +  
labs(subtitle='Comparison of weights against heights of baseball players') +  
xlab('Height') +  
ylab('Weight')  
```


## Question 12. Team Washington
**Obtain a subset of the data with only Team Washington. Using complete sentences, give the number of players in Team Washington.**
```{r, results='asis'}
subset.was <- baseball[baseball[, "Team"] == "WAS", ]
sprintf("There are %d baseball players who play for team Washington (WAS)", nrow(subset.was))
```

## Question 13. Old Team Washington
**Obtain another subset with Team Washington and only players older than 28. How many are there in this veteran group?**
```{r, results='asis'}
subset.was_over_28 <- baseball[baseball[, "Team"] == "WAS" & baseball[, "age"] > 28, ]
sprintf("There are %d baseball players who are over 28 years old and play for team Washington (WAS)", nrow(subset.was_over_28))
```
