---
title: "R - Assignment RMD Basic"
author: "Sample Solution"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r}
# 1. Answer each question using in words/paragraph. 
# 2. DO NOT use comments inside code blocks (like here) to answer anything. Those are for notes between coders/self. They 
# will be ignored, and not counted as answers. 
# 3. Keep the line/option    knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# You can temporarily turn this on/off and use other option while you are working on the homework if it helps. 
# The submitted work should have this option selected instead.
# 4. All charts/graphs/tables should have appropriate captions. 
```

# HW - Module 2 week 03

## Question 1. Import
**Read in the file with all the baseball players height, weight and age info (BaseballHeightWeight.csv) as a dataframe and save it as baseball. Then display the structure of the dataframe.**  

```{r Q1}
# This is coder's comments
# Since my RMD and csv files are in the same working directory, I do not need to give the full path to the csv file. Alternatively, you can use the functions getwd() and setwd() to handle it.

baseball <- data.frame(read.csv("BaseballHeightWeight.csv"))
# baseballdf <- read.csv("BaseballHeightWeight.csv")
# You can compare the two above lines, and look at the object called baseball in the "Environment" on the upper right. 
# Turns out either method gives the same result. In some cases (depending on the .csv file), they might differ. To be safe, I usually use data.frame( ) afterwards.
```

The function `read.csv()` was used to load the dataframe. In addition, we can see that there are `r nrow(baseball)` in the dataframe. 
Note that `read.csv()` is just one of the many functions you can use to import data files. 

The structure is printed below:

```{r, result="asis"}
# need results="asis" to overwrite the default, and show the result block in Html/pdf
str(baseball)
```

I have yet to find a simple and easy way to "pretty-print" such structure in R output. Most of the time we don't need to show these to the readers though. 


## Question 2. Head and Tail
**Print out the first 5 and the last 3 rows of the dataframe.**  

The `head()` and `tail()` functions are what we need.

```{r, results='markup'}
head(baseball, n=5)
tail(baseball, n=3)
```

But we should use the modified version in `ezids` from now on:

```{r, results="asis"}
# you can use "markup" or "asis" here to get the same result.
xkabledplyhead(baseball, 5, "The first 5 rows using `xkabledplyhead` function." )
```
and the last three rows:
```{r, results="asis"}
# you can use "markup" or "asis" here to get the same result.
xkabledplytail(baseball, 3, "The last 3 rows using `xkabledplytail` function." )
```


## Question 3. Header change
**Change the headers/column names from Weight.pounds. to weight, from Height.inches. to height, and Age to age.**  

Done. (No output.)   

```{r}
# You can use names or colnames to change column names like here. Also notice that "=" and "<-" are mostly the same as well.
# colnames(baseball)[4] <- "height"
colnames(baseball)[4:6]=c("height","weight","age")
# The two lines above produces no output. To see the changes, use our head in the next question
```

## Question 4. Head again
**Print out the first two lines again to see the new column names.**  

The first two rows are printed here.  

```{r, results="asis"}
xkabledplyhead(baseball, 2, "Columns renamed." )
```



## Question 5. Summary
**Find the statistics (summary) of the heights, weights, and ages of the players. Using complete sentences, give the reader these summary.**  

The 5-number summary of height by itself (plus the mean) is (use inline R codes here, not just type in the number by hand) `r summary(baseball$height)`. The code-block format doesn't look too nice. You can try that yourself. 
```{r, results='markup'}
summary(baseball$height)
```

I would use the modified function in `ezids` (exclude the name columns): 
```{r, results="asis"}
xkablesummary(baseball[4:6], title = "Baseball players summary of selected statistics" )
```


## Question 6. Subset
**You should find an issue with the weight data with an NA value. Use the `subset( )` function, select that data point (save as shyboy) and find out who the shy boy was.**

Using the `subset()` function, we found the shyboy:
```{r, results="asis"}
shyboy = subset(baseball, is.na(weight))
xkabledplyhead(shyboy, 1, title = "Our shyboy" )
```

## Question 7. Remove Shyboy
**Remove shyboy from baseball, and then show the number of observations of the cleaned up dataframe baseball.**  


```{r Q7, include=F}
baseball = subset(baseball, weight>0) 
```

After removing the data point with invalid weight data, there are (again, use inline R codes here, not just type in the number by hand) `r length(baseball$height)`.


## Question 8. Boxplot
**Make a quick boxplot of the weight of the players. Use the `title( )` function immediately to add a title to the chart. (Try color #0000FF for blue.)**  

```{r Q8, echo=F}
boxplot(baseball$weight, main="Baseball player weight boxplot", ylab="wieght (lbs)", col='#0000FF')
```

A quick glance from the plot suggest the weight distribution is kind of symmetric, other than a few outliers on the right tail.  

We can also get into the habbit of making better plots using the `ggplot2::ggplot()` library/function. If it's some quick plots for our own use, the basic plots are good and quick. The ones from `ggplot2` gives you more control and better visual usually.

```{r}
loadPkg("ggplot2")
ggplot(baseball, aes(y=weight)) + 
  geom_boxplot() + 
  geom_boxplot( colour="orange", fill="#7777cc", outlier.colour="red", outlier.shape=8, outlier.size=4) +
  labs(title="Baseball player weight boxplot using `ggplot`",x="", y = "weight (lbs)")
```

## Question 9. Histogram
**Make a histogram of the weight of the players. Try color 'green' this time. The function understands the word 'green' just fine. Give it a better title than the default. Does the distribution looks normal?**

```{r Q9, echo=F}
hist(baseball$weight, main="Histogram of baseball player weight", col = 'green' ) 
```

The histogram shows somewhat a normal-looking distribution, with a slight right-skew. 

Just like before, we can use `ggplot2::ggplot()` to make this plot:

```{r}
ggplot(data=baseball, aes(weight)) + 
  geom_histogram(breaks=seq(150, 300, by = 10), 
                 col="red", 
                 fill="green", 
                 alpha = .7) + # opacity
  labs(title="Baseball player weight histogram using `ggplot`") +
  labs(x="weight (lbs)", y="Frequency") 
```



## Question 10. Scatterplot
**Use ggplot, make a plot with weights vs heights of the players, color by Teams. Give it an appropriate title.**  

In all scientific literature, a graph is always referred as Y against X (y vs x). 
Make sure you put "weights" on y as requested here.

```{r Q10, echo=TRUE}
# loadPkg("ggplot2")
ggplot(data=baseball)+
  geom_point(mapping = aes(x=height, y=weight, color=Team))+
  ggtitle("Scatter plot of baseball player weigth(y, lbs) vs height(x, inches)")
```

## Question 11. Plot by age
**Make a plot with weights vs heights of the players, color by age.**

I will use `ggplot()` here again. 
```{r Q11, echo=F}
ggplot(data=baseball)+
  geom_point(mapping = aes(x=height, y=weight, color=age))+
  ggtitle("Scatter plot of baseball player weigth(y, lbs) vs height(x, inches)")
```

Notice when colored sorted by a discrete variable `Team` (categorical, qualitative, nominal variable, in R, it is a factor level variable) and numeric `age` (numerical, quantitative, ratio level variable), the plot function handles them differently as you can see from the chart legend.


## Question 12. Team Washington
**Obtain a subset of the data with only Team Washington. Using complete sentences, give the number of players in Team Washington.**

```{r Q12, include=F}
# There are others to subset. Whatever works for you is fine 
teamwash <- subset(baseball,Team=='WAS')
```

We obtained the subset of players with `Team=='WAS'`, which results with `r length(teamwash$weight)` members in the team. 


## Question 13. Old Team Washington
**Obtain another subset with Team Washington and only players older than 28. How many are there in this veteran group?**

```{r}
teamoldwash <- subset(baseball,Team=='WAS' & age > 28)
```

There are `r length(teamoldwash$weight)` in this category.  


