---
title: "Intro to DS - PCA / PCR"
author: ""
date: "today"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```


# USDA Nutrient Dataset  
This data was from USDA National Nutrient dataset, pre-processed by @craigkelly from data.world. 

## Preparation  

### Question 1  
**Import the dataset**  
Import the dataset into R. There should be 45 variables. Take a quick look at the data. Just str, head, and tail will do for now.  
```{r}
data <- read.csv(file='USDANationalNutrient.csv',
                 header=T, sep=',', fill=T, strip.white=T,
                 blank.lines.skip=T)
xkabledplyhead(data)
str(data)
xkabledplytail(data)
```


### Question 2  
**Drop the first 7 variables**  
Save the rest as a dataframe, and call it `nndf1`.   

```{r}
nndf1 <- data[, seq(-7, -1)]
colnames(nndf1)
```

## OLS full model  

### Question 3  
**Issue with full OLS models**  
All the variables are now numerical. Let us build a quick model (full model) to predict `Energy_kcal`, and another full model to predict `VitB12_mcg`. (No fancy output is required.) Vegetarian diet can often be low on B12, which makes one get tired easier. Having a model could be useful. What are the NAs for the coefficients mean? (Try google.) 
```{r}
quick_model_energy <- lm(Energy_kcal~., data=nndf1)
quick_model_vitb12 <- lm(VitB12_mcg~., data=nndf1)

# quick_model_energy
# summary(quick_model_energy)
```

```{r}
# quick_model_vitb12
# summary(quick_model_vitb12)
```

NAs are caused by multicollinearities in the predictor variables. One or more of the vectors in the predictor set can be reproduced by a linear combination of other vectors that have non-trivial solution. For vectors to be linearly independent, the linear combination of them should yield 0 only using a trivial solution (all coefficients of linear combination must be 0).  

### Question 4  
**Fix duplicate/collinear variables**  
You probably realize where the problem comes from. Alright then, let us  

* only keep the first 23 columns, 
* standardize all the variables, and 
* the save the resulting dataframe as `nndf`  

```{r}

nndf <- scale(as.matrix(nndf1[, seq(1, 23)]))
nndf <- data.frame(nndf)
```

Obtain a simple correlation matrix for the dataframe. If you use the function `xkabledply()` on the correlation matrix, it will print out a nice scrollable table for html and your console output.  
```{r}
xkabledply(cor(nndf))
```

### Question 5  
**VitB12_mcg**  
Build a quick full model again for “VitB12_mcg”. What is the $R^2$ value? Notice that using PCA won’t improve the $R^2$. It will only match it if you use all 22 PC-components (1 target, 22 explanatory variables). However, this can be the baseline number to compare.  

```{r}
quick_model_vitb12 <- lm(VitB12_mcg~., data=nndf)
quick_model_vitb12
summary(quick_model_vitb12)
```

The adjusted $R^2$ is `r summary(quick_model_vitb12)$adj.r.squared` is almost the same as Multiple $R^2$. Hence the variables that we had included weren't too bad.  

## PCA  

### Question 6   
**PCA decomposition**  
Instead of PCR, let us run the PCA on this dataset. The dataframe is already centered and scaled. Use  
`nnpcomp <- prcomp(nndf)` to get the 23 PC-components. For `PC1`, which are the four strongest components it consists of?  
(Ans: Riboflavin_mg at -0.3413, Niacin_mg at -0.3378, and so forth.)  
And which are the four strongest for `PC2`?   
```{r}
nncomp <- prcomp(nndf)
nncomp
```
Carb_g at 0.4434, Sugar_g at 0.3588, Protein_g at -0.3434, VitB12_mcg at -0.3550 are the four strong components in PC2.  

### Question 7  
**BiPlots**  
Make a biplot between `PC1` and `PC2` like this: `biplot( nnpcomp, c(1,2), scale=0 )`. Continue to plot `PC1` with `PC3` (use `c(1,3)`), `c(1,4)`, `c(1,5)`, and `c(1,6)`. Also make biplots for `c(2,3)`, `c(3,4)`, `c(4,5)`, `c(5,6)`.     

```{r}
biplot(nncomp, c(1, 2), scale=0)
biplot(nncomp, c(1, 3), scale=0)
biplot(nncomp, c(1, 4), scale=0)
biplot(nncomp, c(1, 5), scale=0)
biplot(nncomp, c(1, 6), scale=0)
biplot(nncomp, c(2, 3), scale=0)
biplot(nncomp, c(3, 4), scale=0)
biplot(nncomp, c(4, 5), scale=0)
biplot(nncomp, c(5, 6), scale=0)
```

### Question 8  
**PVE**  
Does the plots show that `PC1` capture much more variance than `PC2`? `PC6`? (Of course we can confirm these with the PVE plot and calculations if we like. It is not required in this hw here.)     

Yes `PC1` captures more variance than `PC2` and `PC6` as well. This is also evident from seeing the normalized variance in each component.  
In `PC1` vs `PC6` plot, I can see `Sugar_g`, `Carb_g` and `Selenium_mcg` are above 0. Hence, they are more stronger in `PC6` than in `PC1`.  
Likewise, `VitE_mg` and `VitC_mg` are strong components in `PC6`.  


### Question 9  
**Interesting plots**  
If you look at the plot between `PC1` and `PC2`, it looks like a boomerang shape. (See image file in the hw zip.) Recall that `PC1` and `PC2` consists of rather different linear combinations of the original variables. And `PC1`-`PC2` have zero correlation. The chart shows us although `PC1` and `PC2` are overall un-correlated, when `PC1` change, there are only two different trends that `PC2` will change, as if there are only two types of nutrients in the 8000+ observations. Maybe these two kinds of food/items should be analyzed separately with two different models. These are some of the patterns and structures that PCA might be able to reveal.  

At any rate, do you see other charts showing unexpected patterns?
     
## Optional  

We will skip the model building using PCR here as it turns out not a productive exercise in that direction. While we are having this huge dataset, with 23 variables, this is actually a good candidate for performing the feature selection using leaps.  

If you just look at the dataset and guess what are the good predictors to model Energy_kcal, what would you pick?  

Now try run it by leaps regsubsets. What is the minimum number of predictors you will need, if you want to have a 90% $R^2$?  


```{r, include=FALSE}
# unloadPkg(???)
```



