---
title: "Chi-squared  test"
author: "Edwin Lo, GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```


# $\chi^2$ Sqaured

Again, we are not focusing on the neatness and styling of the output here. Just how to run the tests and understand the results.  

## Goodness of Fit test

Now try $\chi^2$ Goodness of Fit test. We need the observed frequencies, and the expected frequencies. Run the $\chi^2$ test, and quickly look at the test summary: 
```{r chisq_goodness1}
grades = c(A=837, B=861, C=690)
exp_freq = c(.36, .35, .29)
# Next perform chi-square test
chigoodness = chisq.test(grades, p = exp_freq) # p stands for proportions
# look at the result
chigoodness
```

The resulting 'object' is complex. You can inspect this `chigoodness` object in the `Environment` tab. The names or indices is shown here:  
```{r, results="markup"}
names(chigoodness)
```


We can extract the $\chi^2$ value of the test from `chigoodness$statistic` = `r chigoodness$statistic`, while extracting the p-value from `chigoodness$p.value` = `r chigoodness$p.value`. The degree of freedom is given by `chigoodness$parameter` = `r chigoodness$parameter`.

The expected values of the frequencies is given by `chigoodness$expected`. Since this part is a simply array object in R, we can use our `xkabledply()` function to neatly print it out.

```{r}
xkabledply( as.table(chigoodness$expected), title = "Expected frequencies", wide = T)
```


```{r, eval=F}
# This block is not shown, but we can run/execute this chunk to see the results easily when coding
chigoodness$statistic
chigoodness$parameter
chigoodness$expected
```


And we can also calculate the p-value directly like this: `pchisq(1.3673, 2, lower.tail=F)` = `r pchisq(1.3673, 2, lower.tail=F)`. We always use the right tail for goodness of fit test. 

```{r, eval=F}
# This block is not shown, but we can run/execute this chunk to see the results easily when coding
pchisq(1.3673, 2, lower.tail=F) # always right tail for test of independence as well
```


## Test of Independence  

Next, let's try using Chi-squared test of independence. We need a "contingency table". Before that, similar to what we did in the ANOVA example, we better re-order the ordinal level variables the proper way.  

```{r chisq_test_indep, results="markup"}
loadPkg("MASS")
surveySmokeExer = MASS::survey 
surveySmokeExer$Smoke <- factor(surveySmokeExer$Smoke, order=T, levels = c("Never","Occas","Regul","Heavy"))
surveySmokeExer$Exer <- factor(surveySmokeExer$Exer, order=T, levels = c("None","Some","Freq"))

contable = table(surveySmokeExer$Smoke, surveySmokeExer$Exer)
# contable # the contingency table
xkabledply(contable, title="Contingency table for Smoke (habit) vs Exercise (level) in library 'MASS'")
```

Next, we can perform the $\chi^2$ test, and take a quick look at the summary result:

```{r, results="markup"}
chitest = chisq.test(contable)
chitest
```

The resulting 'object' is complex. You can inspect this `chitest` object in the `Environment` tab. The names or indices is shown here:  
```{r}
names(chitest)
```


```{r, eval=F}
# This block is not shown, but we can run/execute this chunk to see the results easily when coding
chitest$statistic
chitest$p.value
chitest$parameter
```

The $\chi^2$ value of the test is given by `chitest$statistic` = `r chitest$statistic`, while the p-value is `chitest$p.value` = `r chitest$p.value`. The degree of freedom is given by `chitest$parameter` = `r chitest$parameter`.

The expected values of the frequencies is given by `chitest$expected`. Since this part is a table object in R, we can use our `xkabledply()` function to neatly print it out.

```{r}
xkabledply(chitest$expected, title = "Cross table for the expected frequencies between Smoke (habit) vs Exercise (level) in library 'MASS'")
```

And we can also calculate the p-value directly like this: `pchisq(5.4885, 6, lower.tail=F)` = `r pchisq(5.4885, 6, lower.tail=F)`. We alwas use the right tail for test of independence as well. 

```{r, eval=F}
# This block is not shown, but we can run/execute this chunk to see the results easily when coding
pchisq(5.4885, 6, lower.tail=F) # always right tail for test of independence as well
```





