---
title: "Hypothesis Testing"
author: "Edwin Lo, GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Hypothesis Testing

We performed estimations (Z-intervals and T-intervals) in our last exercise. This time, we are performing hypothesis testings.    

## Preparation
Continuing from last time where we found the Z- and T- intervals... 

```{r}
getwd()
# mlb <- data.frame(read.csv("BaseballHeightWeight.csv", header = TRUE))
mlb <- read.csv("BaseballHeightWeight.csv", header = TRUE)
str(mlb)
head(mlb)
colnames(mlb)[4:6]=c("height","weight","age")
# The line above produces no output. To see the changes, use str() or head()
```

```{r results='markup'}
xkabledplyhead(mlb)
# xkabledplytail(mlb, 3)
```


Again, first take note of the mean height and weight for the entire mlb population is height: `r format(mean(mlb$height), digits=5)` inches, weight: `r format(mean(mlb$weight, na.rm = TRUE), digits=5)` lbs. 
Then redo what we had last time with Z- and T-intervals, but put in the appropriate null hypothesis for the Z- and T-tests. 

```{r}
set.seed(123) # just so that everyone have the same sample for comparison
mlbsample = mlb[ sample(nrow(mlb),30), ]
str(mlbsample)
head(mlbsample)
format(mean(mlbsample$height), digits=4)
format(mean(mlbsample$weight, na.rm = TRUE), digits=5)
```

## Z-Test

So let's say we want to test is the mean height is 72.1 inches like last year as our null hypothesis $H_0: \mu = 72.1$ inches, which was the previous year's average. (Notice the use of latex inline formatting of mathematical typesetting, enclosed with a pair of dollar sign \$ \$ for the inline R-code.)

```{r}
loadPkg("BSDA") # for z.test
ztest95_2tail = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31) # 
ztest95_2tail
ztest95_right = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31, alternative = "greater") # 
ztest95_right
# ztest99 = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31, conf.level=0.99 )
# ztest99
# ztest50 = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31, conf.level=0.50 )
# ztest50
```

```{r results='markup'}
# force this one result to always show, regardless of global settings
ztest95_right
```

You can also try and see it for yourself, the optional argument for `conf.level=0.50` for example does not change the outcome for the p-value or the conclusion of the test. It only changes the confidence interval of the outcome.

In cases that you want to determine the confidence Z-interval (estimation) and Z-test as well, you can do so by supplying the `conf.level` value in the function, and get both results in one step. Typically, you would want to set the conf.level for the interval to be 1 minus the $\alpha$ level for the test to obtain the appropriate corresponding results. 

Now let us try the left- and right-tailed tests
```{r z_test3}
ztestrighttail = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31, alternative = 'greater' )
ztestrighttail  # p-value is small, reject null, adopt alternative mu is greater than 73.2
ztest99lefttail = z.test(x=mlbsample$height, mu=72.1, sigma.x = 2.31, alternative = 'less' )
# ztest99lefttail  # p-value is large, fail to reject null, which is mu = 73.2 (or greater.)
unloadPkg("BSDA")
```

```{r results='markup'}
# force this one result to always show, regardless of global settings
ztest99lefttail
```

In these cases, the Z-intervals produced are one-sided intervals, which we did not used before. But it is possible to create estimates of one-sided intervals.

## T-Test
Similar for t-test, except it is **easier**, since we do not need to know the standard deviation (sigma) of 
the population. 
```{r}
# t.test is included in the basic R package 'stats'
ttest95 = t.test(x=mlbsample$height, mu=72.1) # default conf.level = 0.95
# t.test(x=mlb$height, mu=73.2, sigma.x = 2.18, conf.level=0.99 )
ttest95
ttest99 = t.test(x=mlbsample$height, mu=72.1, conf.level=0.99 )
# ttest99
names(ttest99)
ttest99$conf.int
ttest99$alternative
ttest99$estimate
```

Again, we can check that the optional argument of `conf.level` here only affects the confidence interval result, but not the p-value.

```{r results='markup'}
# force this one result to always show, regardless of global settings
ttest99
```

And here is the left- and right-tailed cases:  

```{r}
ttestrighttail = t.test(x=mlbsample$height, mu=72.1, alternative = 'greater' ) # conf.level=0.99, 
ttestrighttail  # p-value is small, reject null, adopt alternative mu is greater than 73.2
ttestlefttail = t.test(x=mlbsample$height, mu=72.1, alternative = 'less' ) # conf.level=0.99, 
# ttestlefttail  # p-value is large, fail to reject null, which is mu = 73.2 (or greater.)
```

```{r results='markup'}
# force this one result to always show, regardless of global settings
ttestlefttail
```

I have not found a good way to display these results in better format. For now, we'll just screen dump these onto the html. 

## Two-sample t-test

Let us now try to use the two-sample t-test, which is a similar way of using the t-test, but on two different samples. First, create a different sample.

```{r}
set.seed(999) # just so that everyone have the same sample for comparison
mlbsample2 = mlb[ sample(nrow(mlb),20), ]
# compare the two samples
format(mean(mlbsample2$height), digits=4)
format(mean(mlbsample2$weight, na.rm = TRUE), digits=5)
format(mean(mlbsample$height), digits=4)
format(mean(mlbsample$weight, na.rm = TRUE), digits=5)
```

Next, perform the 2-sample t-test on height and weight
```{r}
ttest2sample_height = t.test(mlbsample$height,mlbsample2$height)
ttest2sample_height
ttest2sample_weight = t.test(mlbsample$weight,mlbsample2$weight)
ttest2sample_weight
```

Notice that the two samples do not need to have the same sample size. Even at level of significance alpha set at = 0.50, we still fail to reject the null. The null hypothesis is that the two samples carry the same average.  

Imagine if we pull the two samples one from NBA players and one from WNBA players, do you expect a high or low p-value for the two-sample t-test?

Now, you can try all the above for the weight data for practice.


