---
title: "Intro to DS - Linear Model part I"
author: "Edwin Lo, GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# HW assignment

## Linear model - quantitative regressors 

### Question 1  
**Import the data, call it `bikeorig.`**  
The `Date` variable is probably imported as factor level variable. In any case, let us remove `Date`, `Casual.Users`, 
and `Registered.Users` in the dataset and save it as a new datafame, call it `bike`. How many variables are in `bike`? 
How many of them are imported as `int`? Feel free to rename longer variable names into shorter ones for convenience.  



### Question 2    
**Select only the subset with `Hour` equal 16 only. Call it `bike16`**  
These are the afternoon rush hour data. How many observations are there?   


### Question 3  
**Before building any models, we should make sure the variables are set up properly.**  
(This problem is solved for you. Codes are given below.)  
Which ones should be recorded as categorical? Convert them now before we proceed to the model building.  

Note: After converting, the correlation function `cor()` will not work with categorical/factor variables. 
I would keep the original `bike16` dataframe as numeric, and use that to 
find the correlation matrix. 
Although technically correlation between categorical and numeric variables are not 
well defined in general, we can still get some useful information if the 
categorical variable is at least at ordinal level. See future discussion 
on using "Pearson" vs "Spearman" methods for correlation tests. 

While the `cor()` function does not accept categorical variables (and therefore 
we cannot use it for `corrplot()`), the `lattice::pairs()` function does not complain 
about categorical columns. We can still use it to get a visual distribution of 
data values from it.
 

```{r}
bike_final = bike16
bike_final$Season = factor(bike16$Season)
bike_final$Holiday = factor(bike16$Holiday)
bike_final$Day = factor(bike16$Day)
bike_final$Workday = factor(bike16$Workday)
bike_final$Weather = factor(bike16$Weather)
str(bike_final)
```

We decided to convert these variables into categorical (factor):  
`Season`, `Holiday`, `Day`, `Workday`, and `Weather`.  Notice that 
the dataframe `bike16` still has all variables numerical, while the df `bike_final` 
include categorical columns that we just converted. 

### Question 4  
**Make a `pairs()` plot with all the variables (quantitative and qualitative).**  


### Question 5  
**Make a `corrplot()` with only the numerical variables.**  
You can either subset the df with only numerical variables first, then create 
the create the cor-matrix to plot. Or you can create the cor-matrix from 
`bike16`, then select select out the portion of the matrix that you want. 
Use options that shows well the relationships between different variables. 
 


### Question 6   
**By using numerical variables only, build a linear model with 1 independent variable to predict the `Total Users`.**  
Choose the variable with the strongest correlation coefficient. Make some short 
comments on the coefficient values, their p-values, and the multiple R-squared value.  


### Question 7   
**Next, add a second variable to the model.**  
Choose the variable with the next strongest correlation, but avoid using 
obviously collinear variables. When you have the model, check 
the VIF values. If the VIF is higher than 5, discard this model, and try the 
variable with the next strongest correlation until you find one that works 
(ideally with vif’s <5, or if you have to, allow vif up to 10). Again, 
comment on the coefficient values, their p-values, 
and the multiple R-squared value.  


### Question 8  
**We will try one more time as in the previous question, to add a third variable in our model.**  


### Question 9  
**For the 3-variable model you found, find the confidence intervals of the coefficients.**  


### Question 10    
**Use ANOVA to compare the three different models you found.**  
You have found three different models. Use ANOVA test to compare their residuals. What conclusion can you draw?





