---
title: "Distributions - Normal, Binomial, etc"
author: "Edwin Lo, GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
# knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Normal and Student's T-distribution

From class notes, we can use these following functions to calculate different quantities relating to the normal distribution and t-distribution

Try normal distribution first: 

```{r}
pnorm(-1.5) # to find out what percentile you are at when you exam comes in at 1.5 sd below mean
pnorm(55,mean=70,sd=10) # same as above, using raw score instead of z-score.

qnorm(.75) # to find what critical z-value will contain 50% of data points

dnorm(-1.2) # probability density function at -1.2 
dnorm(1.2) # same as above, since symmetric

rnorm(3) # create 3 random numbers using normal dist, with mu=0, sigma=1
```
Now try Stident's t-distribution: 

```{r}
pt(-1.5, df = 29) # to find out what percentile you are at when you exam comes in at 1.5 sd below mean
# unlike above, pt function cannot use raw score instead of t-score.

qt(.75, df = 29) # to find what critical z-value will contain 50% of data points

dt(-1.2, df = 29) # probability density function at -1.2 
dt(1.2, df = 29) # same as above, since symmetric

rt(3, df = 29) # create 3 random numbers using t-dist, df = 29
```


# Binomial Distribution

Binomial distribution comes from a binary choice (Y/N, Head/Tail, 0/1, T/F). 
We have have many trials of the events, and find the cumulative total/frequency. 
You can change the `xtot` to see how the distribution changes and approaches normal distribution.

```{r binomial}
# modify from 
# http://rgraphgallery.blogspot.com/2013/04/rg68-plot-binomial-distribution.html
library(ggplot2)
# library(grid)

xtot = 20
xmin = 0
probs = 0.5 # probability of success

x1  <- xmin : (xtot - xmin)
df <- data.frame(x = x1, y = dbinom(x1, xtot, probs))

charttitle = paste("dbinom( x, ", xtot,  ", ", probs , ")")

plot1 <- ggplot(df, aes(x = x, y = y)) + geom_bar(stat = "identity", col = "pink", fill = "pink") + 
  scale_y_continuous( expand = c(0.01, 0)) + 
    labs(x='x', y='Density', title = charttitle) + theme_bw(16, "serif", ) + 
    theme(plot.title = element_text(size = rel(1.2), vjust = 1.5)
  )

print(plot1)
```

# Multinomial Distribution 
Example: dice

```{r dice_multinom}
ndice = 2
trials = 25 # simulated results
rmultinom( n=trials, size = ndice, prob = rep(1/6,6)) # prob = c(1/6,1/6,1/6,1/6,1/6,1/6)
# quick check on the results
```

Let us re-do the simulation with different N-trials, and save the result for further analysis. 


```{r}
ndice = 2
trials = 1000 # simulated results

data.frame( t( rmultinom( n=trials, size = ndice, prob = rep(1/6,6) ) ) ) -> multinomdf # need to take the transpose
1:6 -> colnames(multinomdf)
```

The distribution of the number 1, 2, ... ,6:  
```{r}
plot2 <- barplot(colSums(multinomdf), xlab = "x", ylab = "Frequency") + 
  title(main="Frequency distribution for rolling dice") +
  theme_bw()
print(plot2)
```
To plot the distribution of the sum of the dice rolls, we can create a new column to keep track of the sum in the df:  

```{r}
multinomdf$sum = multinomdf$"1" * 1 + multinomdf$"2" *2 + multinomdf$"3" * 3 + multinomdf$"4" *4 + multinomdf$"5" * 5 + multinomdf$"6" * 6 

library(ggplot2)
ggplot(data=multinomdf, aes(sum)) + 
  geom_histogram(breaks=seq(1, 12, by = 1), 
                 col="red", 
                 fill="blue", 
                 alpha = .7) + # opacity
  labs(x="sum", y="Frequency", title="Sum of dice rolled") 

```